---
title: "04: Data Descriptions and Visualizations"
author: "Thomas Manke"
categories:
  - markdown
  - hist()
  - cor()
  - boxplot()
---

```{r, child="_setup.qmd"}
```
    


## Descriptive Statistics
```{r descriptive}
summary(iris)
plot(iris$Petal.Length) # a plot at last. Simple, but many arguments: ?plot 
hist(iris$Petal.Length) # and a histogram
```

**Task** hist() can have many arguments. Use help to find out how the histogram can be customized (e.g. number of break points, title, colors). Try some of it.

## Boxplot: a more condensed summary
```{r}
boxplot(iris$Petal.Length)
```

**Task**: The boxplot above is for *all* data. Create a boxplot of petal length for species "setosa" only.
```{r boxplot, echo=FALSE}
boxplot(iris$Petal.Length[iris$Species=="setosa"], xlab="setosa", ylab="Petal Length", col="red")
```

Boxplot understands data frames
```{r}
boxplot(iris) 
```

**Query**: What does the boxplot for Species mean?


## Interlude: Factors = categorical variables

Factors denote a special class of R-objects that can be thought of as categories (here: species). They have a given number of *levels* which are internally represented as integers.
```{r factors}
class(iris$Species)
typeof(iris$Species)

# a contigency table ~> histogram for categorical data
table(iris$Species)  
```

***

Boxplot understands factors in data frames
```{r boxplot_factors}
boxplot( Petal.Length ~ Species, data = iris)
```

**Task**: Use help to add three different colors:
```{r colored_boxplot, echo=FALSE}
# use help to determine how to add colors
cols=c("red","yellow","blue")
boxplot( Petal.Length ~ Species, data = iris, las=2,col=cols)
```


## Correlations
If a data set has many numerical variables we often want to understand their correlations structure
```{r corr}
x=iris$Petal.Length
y=iris$Petal.Width
plot(x,y)                              # again: this can be customized
abline(v=mean(x),h=mean(y),col="red")  # add vertical/horizontal lines
cor(x,y)                               # a correlation coefficient: which one?
```

## All-Against-All Correlations
**Task**: remove the Species variable from "iris" and store the result in a new data.frame "niris"
```{r niris, echo=FALSE}
niris=iris[,-5]  # generate new data frame without species variable
str(niris)
```

```{r all_against_all}
cor(niris)  # correlation matrix. Which correlation coefficient?
plot(niris) # provide a visualization for data frames, try also pairs(iris)

# assign species-colors to each observation 
cols = iris$Species                        # understand how color is defined
plot(niris, col=cols, lower.panel=NULL)   # "cols" was defined in task above
```


***
## Review
* R markdown files extend scripts and can be converted to html/pdf
* summary for descriptive statistics
* simple plots with simple commands: plot(), hist(), ...
* lots of parameters for customization: line width, colours, ...
* correlation between variables: cor(), plot() - class specific behaviour
* Notice that the data used was extremely clean and structured: data()
