---
title: "05: Data Modeling"
author: "Thomas Manke"
categories:
  - linear model
  - predictions
  - diagnostics
---

```{r, child="_setup.qmd"}
```

# Recap: All-Against-All Correlations

**Task**: Generate all-against-all correlation plot. Understand:

- `cols`
- selection/unselection of elements `[,-5]`
- plot() function and arguments

```{r cor_plot}
# assign species-colors to each observation 
cols = iris$Species                            # understand how color is defined
plot(iris[,-5], col=cols, lower.panel=NULL)   # "cols" was defined in task above
```

***

# From Correlations to Models

**Goal:** 

Model some dependent variable y as function of other explanatory variables x (features)

$$
y = f(\theta, x) = \theta_1 x +  \theta_0
$$

For $N$ data points, choose parameters $\theta$ by ordinary least squares: 

$$
RSS=\sum_{i=1}^{N} (y_i - f(\theta, x_i))^2 \to min
$$

Easy in R:
```{r ols}
plot(Petal.Width ~ Petal.Length, data=iris, col=Species) # use model ("formula") notation
fit=lm(Petal.Width ~ Petal.Length, data=iris)       # fit a linear model
abline(fit, lwd=3, lty=2)                           # add regression line
```

**Query**: What class is the object `fit`?

**Task**: Extract the coefficients of the fitted line.

```{r get_coef, echo=FALSE}
fit$coefficients
coef(fit) 
```

There are many more methods to access information for the `lm` class
```{r class_methods}
methods(class='lm')
```


# Reporting the fit (model)
```{r lm_summary}
summary(fit)        # summary() behaves differently for fit objects
coefficients(fit)   # more functions for specific elements
confint(fit)        # Try to change the confidence level: ?confint
```

This is a good fit as suggested by a 

- small residual standard error
- a large coefficient of variation $R^2$ 
- large F-statistics $\to$ small p-value 
- and by visualization

Fraction of variation explained by model:
$$
R^2 = 1 - \frac{RSS}{TSS} = 1 - \frac{\sum_i(y_i - y(\theta,x_i))^2}{\sum_i(y_i-\bar{y})^2}
$$



# Predictions (with confidence intervals)

Uncertainties in parameters become uncertainties in fits:
```{r predictions}
x=iris$Petal.Length                       # explanatory variable from fit (here:Petal.Length)
xn=seq(min(x), max(x), length.out = 100)  # define range of new explanatory variables
ndf=data.frame(Petal.Length=xn)           # put them into new data frame

p=predict(fit, ndf, interval = 'confidence' , level = 0.95)
plot(Petal.Width ~ Petal.Length, data=iris, col=Species)
lines(xn, p[,"lwr"] )
lines(xn, p[,"upr"] )

#some fancy filling
polygon(c(rev(xn), xn), c(rev(p[ ,"upr"]), p[ ,"lwr"]), col = rgb(1,0,0,0.5), border = NA)

## using ggplot2 - full introduction later
#library(ggplot2)
#g = ggplot(iris, aes(Petal.Length, Petal.Width, colour=Species))
#g + geom_point() + geom_smooth(method="lm", se=TRUE, color="red") + geom_smooth(method="loess", colour="blue")
```

# Poor Model

Just replace "Petal" with "Sepal"
```{r poor_model}
plot(Sepal.Width ~ Sepal.Length, data=iris, col=cols)  
fit1=lm(Sepal.Width ~ Sepal.Length, data=iris)     
abline(fit1, lwd=3, lty=2)    
confint(fit1)                     # estimated slope is indistinguishable from zero
summary(fit1)
```
*Interpretation*: 

- slope is not significantly distinct from 0.
- model does not account for much of the observed variation.


**Task**: Use the above template to make predictions for the new poor model.
```{r poor_model_pred, echo=FALSE}
x=iris$Sepal.Length                       # explanatory variable from fit (here:Sepal.Length)
xn=seq(min(x), max(x), length.out = 100)  # define range of new explanatory variables
ndf=data.frame(Sepal.Length=xn)           # put them into data frame

p=predict(fit1, ndf, interval = 'confidence' , level = 0.95)  # predict values

plot(Sepal.Width ~ Sepal.Length, data=iris, col=Species)
lines(xn, p[,"lwr"] )
lines(xn, p[,"upr"] )
```


# Factorial variables as predictors
In the iris example the "Species" variable is a factorial (categorical) variable with 3 levels.
Other typical examples: different experimental conditions or treatments.

```{r model_factorial}
plot(Petal.Width ~ Species, data=iris)
fit=lm(Petal.Width ~ Species, data=iris)
summary(fit)
```
*Interpretation*:

- "setosa"     (1st species) has mean Petal.Width=0.246(29) - reference baseline
- "versicolor" (2nd species) has mean Petal.Width = Petal.Width(setosa) + 1.08(4) 
- "virginica"  (3rd species) has mean Petal.Width = Petal.Width(setosa) + 1.78(4)

# Anova
`summary(fit)` contains information on the individual coefficients. They are difficult to interpret

```{r anova}
anova(fit)    
```

**Interpretation**: variable "Species" accounts for much variation in "Petal.Width"

***

# More complicated models
Determine residual standard error `sigma` for different fits with various complexity

```{r model_comp}
# A list of formulae
formula_list = list(
  Petal.Width ~ Petal.Length,                 # as before (single variable)
  Petal.Width ~ Petal.Length + Sepal.Length,  # function of more than one variable
  Petal.Width ~ Species,                      # function of categorical variables
  Petal.Width ~ .                             # function of all other variable (numerical and categorical)
)

sig=c()
for (f in formula_list) {
  fit = lm(f, data=iris)
  sig = c(sig, sigma(fit))
  print(paste(sigma(fit), format(f)))
}

# more concise loop using lapply/sapply
# sig = sapply(lapply(formula_list, lm, data=iris), sigma)

par(mar=c(4,20,2,2))
barplot(sig ~ format(formula_list), horiz=TRUE, las=2, ylab="", xlab="sigma")
```

... more complex models tend to have smaller residual standard error (overfitting?)

***

# Model Checking: Diagnostic Plots
"fit" is a large object of the lm-class which contains also lots of diagnostic informmation. 
Notice how the behaviour of "plot" changes.
```{r fit_diag}
fit=lm(Petal.Width ~ ., data=iris)
op=par(no.readonly=TRUE)   # safe only resettable graphical parameters, avoids many warnings
par(mfrow=c(2,2))          # change graphical parameters: 2x2 images on device
plot(fit,col=iris$Species) # four plots rather than one
par(op)                    # reset graphical parameters
```
more examples here: http://www.statmethods.net/stats/regression.html

Linear models $y_i=\theta_0 + \theta_1  x_i + \epsilon_i$ make certain assumptions ($\epsilon_i \propto N(0,\sigma^2)$)

* residuals $\epsilon_i$ are independent from each other (non-linear patterns?)
* residuals are normally distributed
* have equal variance $\sigma^2$ ("homoscedasticity")
* no outliers (large residuals) or observations with strong influence on fit

***

# Review
* dependencies between variable can often be modeled
* linear model lm(): fitting, summary and interpretation
* correlation coefficients can be misleading
* linear models may not be appropriate $\to$ example(anscombe)

