---
title: "01: First Steps"
author: "Thomas Manke"
date: "`r date() `"
output: html_document
---

Familiarize yourself with the layout of the various panels in Rstudio. The most important window is the so-called "Console" with the prompt (">"). This is where much of the interaction with the R-software will take place. The other windows are for: output, help and other information.


# A pocket calculator
In the Console try to repeat and understand the following operations
```{r}
2+2
2^3      # This is a comment: try also 2**3
1+2+3/6  # beware of precedences
sin(pi/2)  # more complicated functions and built-in constants
sqrt(2)
factorial(20) # some big numbers. Notation?
sqrt(-1)
sqrt(-1 + 0i)  # complex numbers
```

Euler Equation: $1+e^{-i\pi}=0$

```{r}
1+exp(-1i*pi) # Why is this not zero?
```

## History
All R-commands are tracked in a *history*. It maybe accessed using the cursors (or the history panel in RStudio). This is extremely useful to navigate to previous commands, repeat them, or change them slightly.

**Task**: Using the history, determine $\sqrt 3$.

## First help
There are various different ways to get help
```{r, echo=TRUE}
?exp
??exp
# example(exp)  # later
```

Use help to find out more about: sin, or factorial, or anything else of interest.

## Community help and References
* online courses: https://www.datacamp.com, http://software-carpentry.org/lessons
* bootcamp: http://www.r-bloggers.com/r-bootcamp-materials  http://jaredknowles.com/r-bootcamp 
* software carpentry: http://swcarpentry.github.io/r-novice-gapminder
* http://stackoverflow.com, http://www.r-bloggers.com/
* QuickR: http://www.statmethods.net 
* Used here: https://github.com/jknowles/r_tutorial_ed/


When seeking help from the community, it is important to also send the output of

```{r}
sessionInfo()
```

****

# Assigning values to objects
Goal: store results of computation in new variables
```{r}
x <- 2*pi
2*pi -> x

x      # see value of object x = show(x) 

# What is the value of object "X"?

x < -2 # Careful: what's going on here?
x

x=2 # similar in most other languages. Notice that x is overwritten
x
x==3  # annother logical comparison
y=x+x # passing objects to functions (+)


# defining character objects
u="a" 
v="b" 


# adding character objects
w=paste(u,v) # more options? Try ?paste
```

****

# Simple Vectors
Vectors illustrate how complex data structures can be built from smaller blocks. Here we learn how to create and inspect vectors.
```{r}
v=c(1,2)            # combine arguments into vector
v                   # display v
str(v)              # inspect structure of  v (especially useful for long and structured objects)
typeof(v)           # inspect type of v ( ~ storage mode)
class(v)            # inspect class of v (determines how certain functions will work with v)

is.vector(v)        # a function with a logical result: TRUE/FALSE
length(v)
sum(v)
summary(v)
```


```{r}
v=c(1,pi,"hello")  # what happens here?     
str(v)
typeof(v)
summary(v)        # notice the changed behaviour of summary()
```

# Efficient Vector Generation
```{r}
v=c(1,2,3,4,5)    # combining many elements can get cumbersome
v=1:5             # same much easier
v=10:-10          # intuitive

v=seq(from=1, to=5, by=1) # same as before but more flexibilty
v=seq(1,5,0.5)            # short-hand
?seq                      # understand other parameter options for seq()
```

**Task**: Create a vector with 5 numbers equally spaced from +2 to -1.9
```{r, echo=FALSE}
v=seq(2,-1.9,length.out=5)
v
```

**Task**: Understand the difference between the two vectors v1 and v2
```{r, eval=FALSE}
v1=1:10-1
v2=1:(10-1)
```

# Accessing Vectors
Frequently we need to access specific elements from a vector
```{r eval=FALSE}
v=1:5             
v[2]              # extract second element from vector
i=1:3             # define an index vector
v[i]              # accessing multiple elements
v[-i]             # what's this ?
v[5:6]            # Careful: what happens here ?
```


### Vectors elements can be named
This can be useful for easier access to vector element
```{r eval=FALSE}
v=c(Thomas=2, Devon=1)
names(v)
v[1]
v["Devon"]
names(v)[1]="TJ"  # renaming names
v
```


R also knows some *built-in* vectors: e.g. letters
```{r}
letters
```

**Task**: Name the elements of vector $v=1:15$ with the first letters of the alphabet.
```{r echo=FALSE}
v=1:15
names(v)=letters[1:15]
v
```


```{r}
v=-2:2
names(v)=letters[1:4]    # can you see all 4 vectors?
str(v)
names(v)[5]              # what is this ?
names(v)=LETTERS[20:24]  # another built-in vector
```


### Type Conversion
The types of variables and vectors can be changed. Try to understand the following commands
```{r eval=FALSE}
as.character(1:10)
as.logical(LETTERS)
as.logical(c(-1,0,0.1,2,NaN,NA))
```

### Vector Operations
```{r eval=FALSE}
x=c(1,2,3)
y=c(2,2,2)
x+y     # elementwise addition
x*y     # ... multiplication
x^y     # ... exponentiation
x > y   # ... comparisons
x %% 2  # modulo operators: try y %% 2
x %*% y # scalar product

```

### Vectors can be recycled. 
```{r eval=FALSE}
v=1:4
v+1
v+c(10,20)
v+c(10,20,30)  # Notice and understand the warning message
v %*% c(10,20) # Error: why?
```

**Task**: Create a *reverse* vector
```{r, echo=FALSE}
rev(LETTERS)
```

Actually, even single variables are vectors (with length=1)
```{r}
is.vector("a")
is.vector(pi)
is.vector(NaN)
```
****

# Learning Curve:
* work with console and R output
* simple functions (input, output) and logical operation
* numerical precision
* creating simple objects and assigning values
* inspecting objects: v, str(v), typeof(v), lenght(v)
* case-sensitivity
* flexible creation of vectors
* vector indices and subsetting
* naming vector elements
* vector operations and recycling
* undefined values: NaN, NA
* type conversions: numbers->characters->logical
* functions can behave differently on different input: summary()
* see warning messages and errors
* use help to understand about functions: ?exp
