---
title: "03: Data Descriptions and Visualizations"
author: "Thomas Manke"
date:  '`r Sys.Date()`'
output:
  html_document:
    df_print: paged
    code_folding: show
    toc: yes
    toc_depth: 2
---

```{r, child="_setup.Rmd"}
```
    
# R Markdown

R scripts (.R) help to define and run reproducible analysis workflows, but they lack documentation (other than comments).

R Markdown files (*.Rmd) combine scripts with powerful text formatting.
They can be run (="rendered"="knit") to produce html and pdf.


**Task**:
In Rstudio, open a new R markdown document with `File > New File > R Markdown...`.   
This will open a template for an Rmd file that can be knit 

**Notice**:  
- To convert this file into html press `Knit` - try it out! You may have to save it first as - e.g. "first.Rmd"
- There is a `YAML` header that contains parameters which will affect the rendering process - customize them
- In R studio the code blocks can be run individually using the embedded "Play" button - try it out

# Lecture materials
This lecture series is written in R Markdown and is publically available on github: https://github.com/maxplanck-ie/Rintro.git  (!branch: 2023.03!)

You can access this material in various different ways
- download as zip archive and unpack
- Console (advanced): git clone -b 2023.03 https://github.com/maxplanck-ie/Rintro.git
- Rstudio (recommended): open a new Rproject ("New Project > Version Control > Git > ) and provide the same link as Repository URL

# Descriptive Statistics
```{r descriptive}
summary(iris)
plot(iris$Petal.Length) # a plot at last. Simple, but many arguments: ?plot 
hist(iris$Petal.Length) # and a histogram
```

**Task** hist() can have many arguments. Use help to find out how the histogram can be customized (e.g. number of break points, title, colors). Try some of it.

# Boxplot: a more condensed summary
```{r}
boxplot(iris$Petal.Length)
```

**Task**: The boxplot above is for *all* data. Create a boxplot of petal length for species "setosa" only.
```{r boxplot, echo=FALSE}
boxplot(iris$Petal.Length[iris$Species=="setosa"], xlab="setosa", ylab="Petal Length", col="red")
```

Boxplot understands data frames
```{r}
boxplot(iris) # What does the boxplot for Species mean?
```

# Interlude: Factors = categorical variables

Factors denote a special class of R-objects that can be thought of as categories (here: species). They have a given number of *levels* which are internally represented as integers.
```{r factors}
class(iris$Species)
typeof(iris$Species)
ts=table(iris$Species)  # returns a contigency table ~> histogram for categorical data
barplot(ts, col=rainbow(3), ylab="observations", cex.names=0.9)
pie(ts,col=rainbow(3))  # if you really must
```

***

Boxplot understands factors in data frames
```{r boxplot_factors}
boxplot( Petal.Length ~ Species, data = iris, las=2) # what does las=2 do ?
```

**Task**: Use help to add three different colors:
```{r colored_boxplot, echo=FALSE}
# use help to determine how to add colors
cols=c("red","yellow","blue")
boxplot( Petal.Length ~ Species, data = iris, las=2,col=cols)
```


# Correlations
If a data set has many numerical variables we often want to understand their correlations structure
```{r corr}
x=iris$Petal.Length
y=iris$Petal.Width
plot(x,y)                              # again: this can be customized
abline(v=mean(x),h=mean(y),col="red")  # add vertical/horizontal lines
cor(x,y)                               # a correlation coefficient: which one?
```

# All-Against-All Correlations
**Task**: remove the Species variable from "iris" and store the result in a new data.frame "niris"
```{r niris, echo=FALSE}
niris=iris[,-5]  # generate new data frame without species variable
str(niris)
```

```{r all_against_all}
cor(niris)   # correlation matrix. Which correlation coefficient?
pairs(niris) # provide a visualization, try also plot(iris)

# assign species-colors to each observation 
cols = iris$Species                        # understand how color is defined
pairs(niris, col=cols, lower.panel=NULL)   # "cols" was defined in task above
```


***
# Review:
* use and understand frequently used data sets (iris)
* summary  for descriptive statistics
* plot() for X-Y plots and overloading for other classes (lm)
* hist(), boxplot()
* customize arguments: line width, colours, ...
* correlations: cor()
* Notice that the data used was extremely clean and structured: data()
