---
title: "Image Power: Vizualizing Matrices"
author: "Thomas Manke"
date: "31 May 2016"
output: html_document
---


# Image:
Rather than looking at all pairwise correlation coefficients, it is often easier to visualize them as a colour-coded matrix image.
```{r}
niris=iris[,-5]
C=cor(niris)
C
image(C)
```

**Task**: Another famous and pretty example. Who is volcano?
```{r}
image(volcano, axes=FALSE)
```

**Task**:
Using the image() call, try to reproduce the following image for the full iris data (not just its correlation matrix)
```{r, echo=FALSE}
M=as.matrix(niris)
image(M)
```

Be aware:

* Columns (=variables) appear as rows in image(). How could this be changed?
* the axes are meaningless. Default: $0 \ldots 1$
* there is a large difference in scales for the different variables
* there is an implicit color-scheme: which one?

Below we are re-scaling the iris-matrix
```{r}
S=scale(M)            # scales column-wise
br=seq(-5,7,by=0.5)   # set common break points for the histograms below
myblue=rgb(0,0,1,0.5) # define custom-made colour: transparent blue
hist(M[,"Sepal.Width"], breaks = br)
hist(S[,"Sepal.Width"], breaks = br, add=TRUE, col=myblue)      # add histogram to current plot
legend("topright", c("orig","scaled"), fill=c("white", myblue))

op=par(no.readonly=TRUE)
par(mar=c(1,1,3,7)) # set margins to make room for axis labels
image(S,main="Normalized IRIS data", axes=FALSE)                  # as before, suppress axes
axis(4,at=seq(0,1,length.out=ncol(S)), labels=colnames(S), las=2) # Understand this command
par(op)  # reset graphical parameters
```

## Heatmaps
Heatmaps are advanced images in which rows and columns are re-ordered according
to some distance measure (default: Euclidean) and hierarchical clustering method (default: complete) 
```{r}
heatmap(M)
```

Notice that:

* unlike image(), the heatmap is already in the "canonical" orientation
* the heatmap is usually scaled by rows (columns more useful)
* frequently needs many customizations: label size, remove row-names, ...
* there is an implicit colour scheme: "numerical value $\to$ colour"
* there are multiple extensions of this in other R-packages: e.g. gplots::heatmap.2
```{r}
cols <- rainbow(3)[iris$Species]  # understand how color is defined
heatmap(M,scale="col", RowSideColors=cols,labRow=FALSE, cexCol=0.9)
```

# Sending plots to files
In Rstudio, we can export figures from the "Plots" tab. On the console we can define a pdf file as a new device for all subsequent figures. This is usually done only *after* the image is sufficiently optimized
```{r}
pdf("../data/heatmap.pdf") # similar for jpeg, png, ...
heatmap(M,scale="col", RowSideColors=cols,labRow=FALSE, cexCol=0.9)
dev.off()  # close device = pdf file
```


***

# Review:
* use image() and heatmap() to visualize matrices
* customization of plots again: colours, legends
* fallacies of colour-scaling
* controlling figures with par()
* exporting figures as publication-ready files


