---
title: "Image Power: Vizualizing Matrices"
author: "Thomas Manke"
date: "5 Dec 2016"
output: html_document
---


# Image:
Rather than looking at all pairwise correlation coefficients, it is often easier to visualize them as a colour-coded matrix image.
```{r}
niris=iris[,-5]
C=cor(niris)
C
image(C)
```

**Task**:
Using the image() call, try to reproduce the following image for the full iris data (not just its correlation matrix)
```{r, echo=FALSE}
M=as.matrix(niris)
image(M)
```

Be aware:

* Columns (=variables) appear as rows in image(). How could this be changed?
* the axes are meaningless. Default: $0 \ldots 1$
* there is a large difference in scales for the different variables
* there is an implicit color-scheme: which one?

Below we are re-scaling the iris-matrix
```{r}
S=scale(M)            # scales column-wise
br=seq(-5,7,by=0.5)   # set common break points for the histograms below
hist(M[,"Sepal.Width"], breaks = br)
hist(S[,"Sepal.Width"], breaks = br, add=TRUE, col="red")      # add histogram to current plot
legend("topright", c("orig","scaled"), fill=c("white", "red"))

image(S,main="Normalized IRIS data", axes=FALSE)              # as before, suppress axes

#add axes flexibly
#axis(4,at=seq(0,1,length.out=ncol(S)), labels=colnames(S), las=2) # Understand this command
```

## Heatmaps
Heatmaps are advanced images in which rows and columns are re-ordered according
to some distance measure (default: Euclidean) and hierarchical clustering method (default: complete) 
```{r}
heatmap(M)
```

Notice that:

* unlike image(), the heatmap is already in the "canonical" orientation
* the heatmap is usually scaled by rows (columns more useful)
* frequently needs many customizations: label size, remove row-names, ...
* there is an implicit colour scheme: "numerical value $\to$ colour"
* there are multiple extensions of this in other R-packages: e.g. gplots::heatmap.2
```{r}
cols <- rainbow(3)[iris$Species]  # understand how color is defined
heatmap(M,scale="col", RowSideColors=cols,labRow=FALSE, cexCol=0.9)
```

# Sending plots to files
In Rstudio, we can export figures from the "Plots" tab. On the console we can define a pdf file as a new device for all subsequent figures. This is usually done only *after* the image is sufficiently optimized
```{r}
pdf("../data/heatmap.pdf") # similar for jpeg, png, ...
heatmap(M,scale="col", RowSideColors=cols,labRow=FALSE, cexCol=0.9)
dev.off()  # close device = pdf file
```

# Dimensional Reduction: PCA
```{r}
D=as.matrix(iris[,-5])     # numerical data, some operations below require matrices not data frames (%*%)
s=iris[,5]                 # species attributes (factor)
```

# Goals:
* simplify description of data matrix $D$: data reduction & extract most important information
* maximal variance: look for direction in which data shows maximal variation
* minimal error: allow accurate reconstruction of original data 

```{r}
pca=prcomp(D, scale=TRUE)
```

*Task*: What kind of object is pca?

```{r}
R=pca$rotation  # rotation matrix
S=pca$x         # score matrix = rotated scaled data matrix = Ds %*% R
pairs(S, col=s) # 
plot(R[,1:2], main="Rotations")
# covariance matrix of scores is diagonal (by design)
# --> principal components are uncorrelated
CS=cov(S)   # = t(S) %*% S / (nrow(S) - 1)
CS
```

Notice that the higher components do not add much information, so we may as well represent the transformed data in only the first two dimensions:
```{r}
plot(S[,1:2],pch=21, bg=s)  # score-plot
```

A biplot represents both the observations and the old variables in a new coordinate system (up-to-rescaling)
```{r}
biplot(pca, pc.biplot = TRUE) 
```


***

# Review:
* use image() and heatmap() to visualize matrices
* customization of plots again: colours, legends
* fallacies of colour-scaling
* exporting figures as publication-ready files
* dimensional reduction with PCA


