---
title: "Data Managment: Objects and Files"
author: "Thomas Manke"
date: "`r date()`"
output: html_document
---

# Saving objects as Rdat files
In the process of data exploration we often generate many new data objects. Just as with figures, we may also save data objects for later use. 

The following call saves an "image" of all data obejcts in a compressed file format (conventional: *.Rdat)
```{r, eval=FALSE}
save.image(file="04.Rdat") 
```

**Task**: What happens if the "file=" parameter is omitted? Saving all data can be an overkill. Find out how to save only selected or single objects, e.g. S.

# Saving objects as text files
Rdat files are most convenient for future processing with R.
Sometimes we need to save objects in (tab-separated) text files:
```{r, eval=FALSE}
write.table(iris, file="iris.tsv", sep="\t", row.names=FALSE, quote=FALSE)
```

# Tracking objects
In RStudio, explore the "Environment" tab. On the console there are also many convenient commands to track (and delte) objects from memory.
```{r, eval=FALSE}
ls()     # list all obejct
rm(x)    # remove object x
# rm(list=ls()) # Careful: This removes all objects
```

# Navigation: Where am I, where are my files?
In Rstudio, explore the "Files" tab. On the console:
```{r, eval= FALSE}
getwd()         # get the full path of the current working directory
list.files()    # list all files in working directory
setwd("~")      # change working directory to new directory (here: "~" = home)
list.files(pattern=".Rdat") # list all ".Rdat files in working directory  
```

# The theory of I/O
Reading external data into R is easy, but it is crucial to have a very clean and well-defined data structure. 
```{r, eval=TRUE}
# load("04.Rdat")  # easiest and fastest for previously generated Rdat files
iris_f=read.table(file="../data/iris.tsv", sep="\t", header=TRUE)
str(iris_f)
head(iris_f)
```

# The reality of I/O
In reality, many external data is badly formatted, leading to much wasted time - before any analysis. 


### Mock real
```{r, eval=TRUE, error=TRUE}
file="https://raw.githubusercontent.com/maxplanck-ie/Rintro/master/data/GeneList.tsv"  # locate at github https://github.com/maxplanck-ie/Rintro
read.table(file) 
read.table(file, header=TRUE)                                            # watch out for headers
read.table(file, header=TRUE, comment.char = "%")                        # comment lines
read.table(file, header=TRUE, comment.char = "%", sep="\t")              # separators
read.table(file, header=TRUE, comment.char = "%", sep="\t", quote="\"")  # quote character. What is the default?

# read.delim(file, comment.char = "%")                                   # alternative with certain options in place

# Now fill the data frame
D=read.table(file, header=TRUE, comment.char = "%", sep="\t", quote="\"")

is.na(D)  # any NA's?

colSums(is.na(D))
i=which(rowSums(is.na(D[,3:5]))==0)   # keep rows without NA
j=which(colSums(is.na(D))==nrow(D))   # remove columns with all NA
D[i,-j]
```

### Real real
Notice1: The following code fragments took ~2 hours of trial and error. It is kept only for reference. No need to execute

Notice2: R can read from http. For slow network connections, it will be faster to first download the file to local storage and work from there.


```{r, eval=FALSE}
# consider a data file from one of the first large scale ChIP-studies (Rick Young lab: yeast)
# 1. Find their homepage (google) and go to the "Data Download"" section.
# 2. find the Publication by Lee et al. (2002) Transcriptional Regulatory Networks in Saccharomyces cerevisea.
# 3. In the "Download Raw Data" section copy the second link ("binding of regulators to genes (text file)"
# this file contains binding data (p-values and ratios) for more than 6000 yeast genes and 113 yeast transcription factors

chip="http://jura.wi.mit.edu/young_public/regulatory_network/binding_by_gene.tsv"

# first problem: two headers --> skip first --> skip line 1
D=read.table(file=chip, header=TRUE,  skip=1, sep="\t")
nrow(D)  # 511 << 6000 !!!???

# second problem: quote symbols in names (')
cf=count.fields(chip,sep="\t") # many NA --> quoted fields got missing

D=read.table(chip, header=TRUE,  sep="\t", quote="\"", skip=1)
rownames(D)=D[,1]          # ORF-symbols in column 1, keep them as rownames 
ic=c(1:4,seq(6,230,by=2))  # define column indices to be excluded (only keep p-values)
P=D[,-ic]                  # new data-frame of p-values, exclude superfluous columns and ratios
write.table(P,file="../data/chip.tsv", sep="\t", quote=FALSE)  # write p-values to file
```

### Cleaned Real
Let's start with a cleaned-up data set. Now available at "https://raw.githubusercontent.com/maxplanck-ie/Rintro/master/data/chip.tsv"
(locate this *raw* file on https://github.com/maxplanck-ie/Rintro/)

```{r}
fn="https://raw.githubusercontent.com/maxplanck-ie/Rintro/master/data/chip.tsv"  # remote filename
P=read.table(file=fn, header=TRUE, sep="\t")
```

Describe data and sanity checks
```{r, eval=FALSE}
str(P)            # describe the data structure
which(is.na(P))   # make sure everything is properly defined
which(P<0 || P>1) # make sure p-values are all between 0 and 1
```

Now we can start looking at something more interesting
```{r}
# defined boolean matrix of TF-gene interactions (based on p-value threshold)
pt=1e-3
B=P<pt  # result is matrix.

# reduce problem size, heatmap slow
i=rowSums(B)>1   # only genes with sufficient number of TF-binding
j=colSums(B)>40  # only TF with sufficient number of targets

# heatmap needs numeric matrix ==> convert logical to integer 
B=1*B  # equivalent tricks: B[B]=1; mode(B) = "integer", but _not_ as.integer(B)  # strips attributes, i.e. dimensionsas

heatmap(B[i,j], scale="none", col=c("white","black"), labRow=FALSE)
#d3heatmap(B[i,j], scale="none", col=c("white","black"), distfun = function(c) dist(c, method="binary"))
```

Explore relation between FHL1 and RAP1
```{r}
smoothScatter(log(P[,"FHL1"]),log(P[,"RAP1"]),main="smoothScatter of p-values")  # scatterplot of log(p-values) for two TF 
abline(h=log(pt), v=log(pt), col="red")                                          # "arbitrary" thresholds
cor(B[,"FHL1"],B[,"RAP1"])                                                       # Phi-coefficient = Pearson CC for binary variables
```

Notice that some correlations may be undefined because all entries for certain TF (=columns) are zero
```{r}
cor(B[,"ABF1"], B[,"HAA1"]) 
sum(B[,"HAA1"])
```

Find matrix elements with maximal correlations (ignore NA's and ignore diagonal=1)
```{r}
C=cor(B)
m=max(C[upper.tri(C)], na.rm=TRUE)  # maximum of upper-triagonal matrix. Ignore NA for maximum
idx=which(C==m, arr.ind = TRUE)     # return matrix indices. There are two index pairs  because C is symmetric.
idx
i1=idx[1,1]
j1=idx[1,2]

cat(colnames(C)[c(i1,j1)], C[i1,j1])
```

Contigency tables
```{r}
table(B[,"FHL1"])               # 194 genes bound by FHL1
tb=table(B[,"FHL1"],B[,"RAP1"]) # contingency table: 119 genes bound by FHL1 and RAP1
tb

fisher.test(tb)                 # the overlap is highly unexpected
chisq.test(tb)
```

Generalize: Find other groups of TF with common targets 
```{r}
C=cor(B[,j])                             # correlation matrix. Alternative: C=cor(log(P+eps))
heatmap(C,col = rev(grey.colors(100)) ) 
```

### Packages
Heatmaps can summarize overall data, but are often difficult to explore.
Additional tools are needed. Enter the world of *R packages*.

Explore CRAN "https://cran.r-project.org" orGoogle "interactive heatmaps in R"
```{r}
#install.packages("heatmaply")   # only need to install once (but takes some time: ~10min)
library(heatmaply)               # only need once per session. notice messages and warnings
heatmaply(C,col = rev(grey.colors(100)) ) 
```

Investigating overlaps:
```{r}
# Which genes are commonly bound by FHL1 and RAP1?
names(which(B[,"FHL1"]==1 & B[,"RAP1"]==1))

# create a Venn Diagram of target gene overlaps (with package gplots)
library(gplots)
L=list(FHL1=which(B[,"FHL1"]==1),RAP1=which(B[,"RAP1"]==1),MCM1=which(B[,"MCM1"]==1), STE12=which(B[,"STE12"]==1) )
vn=venn(L)

# better than Venn
library(UpSetR)
upset( as.data.frame(B), sets = c("FHL1", "RAP1", "MCM1", "STE12"), empty.intersections = TRUE )
```

***

# Review:

* exporting objects as Rdat (save.image) or text-file (write.table)
* navigate in directories, know where you are working: ls
* I/O pain: keep data clean and well structured !
* data retyping: data.frame --> Boolean matrix --> Integer
* data exploration and visualization heatmap(), smoothScatter()
* quantitative data descriptions: cor(), table()
* tests: fisher.test(), chi2.test()
* packages: heatmaply, gplots (venn), UpSetR

