---
title: "Data Managment: Objects and Files"
author: "Thomas Manke"
date: "31 May 2016"
output: html_document
---

# Saving objects as Rdat files
In the process of data exploration we often generate many new data objects. Just as with figures, we may also save data objects for later use. 

The following call saves an "image" of all data obejcts in a compressed file format (conventional: *.Rdat)
```{r, eval=FALSE}
save.image(file="04.Rdat") 
```

**Task**: What happens if the "file=" parameter is omitted? Saving all data can be an overkill. Find out how to save only selected or single objects, e.g. S.

# Saving objects as text files
Rdat files are most convenient for future processing with R.
Sometimes we need to save objects in (tab-separated) text files:
```{r, eval=FALSE}
write.table(iris, file="iris.tsv", sep="\t", row.names=FALSE, quote=FALSE)
```

# Tracking objects
In RStudio, explore the "Environment" tab. On the console there are also many convenient commands to track (and delte) objects from memory.
```{r, eval=FALSE}
ls()     # list all obejct
rm(x)    # remove object x
# rm(list=ls()) # Careful: This removes all objects
```

# Navigation: Where am I, where are my files?
In Rstudio, explore the "Files" tab. On the console:
```{r, eval= FALSE}
getwd()         # get the full path of the current working directory
list.files()    # list all files in working directory
setwd("~")      # change working directory to new directory (here: "~" = home)
list.files(pattern=".Rdat") # list all ".Rdat files in working directory  
```

# The theory of I/O
Reading external data into R is easy, but it is crucial to have a very clean and well-defined data structure. 
```{r, eval=TRUE}
# load("04.Rdat")  # easiest and fastest for previously generated Rdat files
iris_f=read.table(file="../data/iris.tsv", sep="\t", header=TRUE)
str(iris_f)
head(iris_f)
```

# The reality of I/O
In reality, many external data is badly formatted, leading to much wasted time - before analysis. The following code fragments took ~2 hours of trial and error
```{r}
# consider file from one of the first large scale ChIP-studies (R.Young lab: yeast)
# this file contains more the binding data (p-values and ratios) for more than 
# 6000 yeast genes and 113 yeast transcription factors
chip="http://jura.wi.mit.edu/young_public/regulatory_network/binding_by_gene.tsv"

# first problem: quote symbols in names (')
D=read.table(chip, header=TRUE,  skip=1, sep="\t")
nrow(D)  # 511 << 6000 !!!???
cf=count.fields(chip,sep="\t") # many NA --> quoted fields got missing
                
# second problem: two headers. need information from both
D=read.table(chip, header=TRUE,  sep="\t", quote="\"", skip=1)
H=read.table(chip, header=FALSE, sep="\t", quote="\"", nrows=1, stringsAsFactors = FALSE)
colnames(D)[1:4]=H[1,1:4]
rownames(D)=D[,1]
ic=c(1:4,seq(6,230,by=2))
P=D[,-ic]         # data-frame of p-values, exclude superfluous columns and ratios
which(is.na(P))   # make sure everything is properly defined
which(P<0 || P>1) # make sure p-values are all between 0 and 1
```

# Appendix
Only now we can start looking at something more interesting
```{r}
# defined boolean matrix of TF-gene interactions (based on p-value threshold)
B=P<1e-3  

Bn=B[rowSums(B)>2, colSums(B)>40] # reduce the problem size
Bn[Bn]=1                          # boolean -> numeric
heatmap(Bn,scale="none",labRow=FALSE, cexCol=0.6)

# compare binding p-values for two TF 
smoothScatter(log(P[,"FHL1"]),log(P[,"RAP1"]),main="smoothScatter of p-values")

table(B[,"FHL1"])               # 194 genes bound by FHL1
tb=table(B[,"FHL1"],B[,"RAP1"]) # contingency table: 119 genes bound by FHL1 and RAP1
tb

# Which genes are bound by FHL1 and RAP1?
common_genes=names(which(B[,"FHL1"] & B[,"RAP1"]))

# fourfoldplot(tb)
# mosaicplot(tb, color = TRUE)
fisher.test(tb)                 # the overlap is highly unexpected
chisq.test(tb)


# create a Venn Diagram of target gene overlaps (with package gplots)
library(gplots)
L=list(FHL1=which(B[,"FHL1"]),RAP1=which(B[,"RAP1"]),MCM1=which(B[,"MCM1"]), STE12=which(B[,"STE12"]) )
vn=venn(L)
```

***

# Review:

* exporting objects as Rdat or text-file
* navigate in directories, know where you are working
* keep data clean and well structured !
* appendix: a (semi-)real-life example
* smoothScatter(), contigency tables, tests, venn (package)

