---
title: "Sampling: Monty Hall Problem"
author: "T. Manke"
date: "24 Feb 2016"
output: html_document
css: ../css/test.css
---

# The Game:
A showmaster presents three closed doors (1,2,3) to a candidate. Only behind one random door is a price (i.e a car).
The canddiate is asked to pick one of the doors. Afterwards, the showmaster opens one of the other two doors which does not contain the price.
Now two doors are still closed and the candidate has the choice to *either stick* to the original choice, *or switch* to the other closed door.
Question: which strategy is more likely to succeed in winning the price?


# Simulation in R
Define the three doors, only one of which hides a price - the winning door "W". The other two doors are loosers (L1 and L2)
```{r}
doors=c("W","L1","L2")
```

First assume that the doors are shuffled randomly
```{r}
doors=sample(doors) # Why does this amount to shuffling?
```

Now the candidate picks a random door "pick" (hoping that it might be the winning door)
```{r}
pick=sample(doors,1)
pick
```

Only the show master knows the winning door. Depending on the picked door, the show master will reveal one door that does _not_ contain the price, and _not_ the picked door. 
```{r}
cs=which(doors != "W" & doors != pick) # depending on pick this could be one or two elements
open=sample(doors[cs],1)
open
```
Notice that the showmaster does not really have a choice when the candidate has picked a loosing door (2/3 of the time).
Now the candidate has the choice to switch to the other remaining door (not open, not previously picked).
The question is: Will the switching strategy be more successful?
```{r}
cc=which(doors != open & doors != pick)
swit=sample(doors[cc],1)
swit
```

With three doors, the candidate has only two choices to stay with "pick" or to switch to "swit"

Now let's play this game N times
```{r}
N=10000
success_stay=0
success_swit=0
for (i in 1:N) {
  doors=sample(doors)

  pick=sample(doors,1)

  open=sample(doors[doors!="W" & doors != pick],1)
  swit=sample(doors[doors!=open & doors !=pick],1)

 # cat(pick, " " , swit, " ", open,"\n")
  
  if (pick=="W") { success_stay=success_stay+1 }
  if (swit=="W") { success_swit=success_swit+1 }
}

cat("N= ",N, "successes with stay= ", success_stay, " successes with switching= ",success_swit,"\n")
```

**Task**: save the above lines in a file called Monty.R.


Congratulations! You have just written your first "R script".
You can share it easily with all people -- especially those who don't like to switch.

# Bayesian Treatment:
If you prefer paper and pencil, this is how to solve this problem analytically.

* **prior probailities**: Pr(1=car)=Pr(2=car)=Pr(3=car)=$\frac{1}{3}$

Assume that the candidate picks door 1 and the showmaster opens door 2 (up to relabeling). Now the candidate would like to know

* **posterior probability**: Pr(3=car|2=shown)= ?

* **Bayes Theorem** (on updating probabilities): Pr(A,B)=Pr(A|B)Pr(B)=Pr(B|A)Pr(A)

* Pr(3=car|2=shown)=$\frac{\mbox{Pr(2=shown|3=car)Pr(3=car)} }{\mbox{Pr(2=shown)}}$

* **law of total probability (marginalization):**
* Pr(2=shown)=Pr(2=shown|1=car)Pr(1=car) + Pr(2=shown|2=car)Pr(2=car) + Pr(2=shown|3=car)Pr(3=car) 
* Pr(2=shown)=$\frac{1}{2} \cdot \frac{1}{3}   + 0 \cdot \frac{1}{3} + 1 \cdot \frac{1}{3} = \frac{1}{2}$
* Pr(3=car|2=shown)=$\frac{ 1 \cdot \frac{1}{3} }{ \frac{1}{2} } = \frac{2}{3}$
* Pr(1=car|2=shown)= 1 - Pr(2=car|2=shown)-Pr(3=car|2=shown)= 1 - 0 - $\frac{2}{3}=\frac{1}{3}$




<div class="important blue">
<br>
Pr(3=car|2=shown) = 2 $\times$ Pr(1=car|2=shown)
<br> <br>
</div>

# Did you understand?
* solving problems (in R) with simulations
* loops (for) and conditionals (if)
* which()
* selection/exclusion in vector
* sample()
* writing and sharing R-scripts.
* prior and posterior (conditional) probabilities
* Bayes Theorem and updating probabilities 
