---
title: "Simulated Data: The Birthday Problem"
author: "T. Manke"
date: "25 Feb 2016"
output: html_document
---

# The Problem:
Given $n$ people at a birthday party, how likely is it that at least two of them share a birthday?

# Sampling Excerices
Sample uniformly (with replacement) from 365 days (no Feb 29)

```{r}
n=40
B=sample(1:365, n, replace=TRUE)  # sample n birthdays
```

Are there any duplicated entries?
```{r}
B
dups=duplicated(B) # What sort of object is dups?
anyDuplicated(B)   # What does this function return? When is it zero?
```

# Simulation in R
With those ingedients, determine the expected fraction of $m=1000$ parties with one or more duplicated birthdays.
```{r}
days=365
m=1000
B=sample(days, m*n, replace=TRUE) # mxn birthdays
dim(B)=c(m,n)                     # Birthday matrix: m parties x n participants
x=apply(B,1,anyDuplicated)        # apply anyDuplicated() function to each row (1) of B
fract=sum(x>0)/m                  # fraction of m parties with duplicates 
fract
```

### Functions
Goal: Convert the above into a reusable piece of software. Write a R-function that returns this fraction of parties.

Parameters: Number of days/year, Number of Participants, Number of parties

Using the history write the following function to a source and save as file (e.g. myFract.R)
```{r}
myFract <- function (n,m=1000,days=365) {
  B=sample(days, m*n, replace=TRUE)  # sample n*m birthdays
  dim(B)=c(m,n)                      # convert to matrix
  x=apply(B,1,anyDuplicated)         # return first duplicated index or zero if none
  fract=sum(x>0)/m                   # fraction of entries in x bigger than zeros
  return(fract)
}
```

To invoke the newly defined function you will need to "source" the file.
Use "source(file)" at the console, or use the appropriate button in Rstudio.

If there are no errors in the script, you should be able to run some tests and examples.

```{r}
myFract(40)
myFract(40,m=100)
```


Now plot the fraction against the size of the party $n \in [1,50]$
```{r}
fract=vector("numeric",50)
for (n in 1:50) { fract[n]=myFract(n) }
plot(1:50, fract, main="Birthday Problem: numerical simulation", xlab="People", ylab="Fraction", 
     type="o", col="red", pch=21, bg="grey",lwd=2)

abline(h=0.5, lty=3, col="blue", lwd=2)
```

With $n>23$ people we expect more than half of all parties to have matching birthdays.

# Analytical treatment
For any given pair of people the probability of success (sharing a birthday) is small $p=1/365 \approx 0.0027$, but in the party setting 
there are many such comparisons $N=n(n-1)/2 \propto n^2$. The total number of successes can be modelled by the **binomial distribution** with success probability $p$ and $N$ number of trials.
Rather than simulating the process (above), we might have also sampled $m$ random numbers from a binomial distribution
```{r}
m=100
n=24
p=1/365
x=rbinom(m,size=n*(n-1)/2, prob=p) # generate m random numbers = number of successes

#  distribution of number of successes
table(x)
barplot(table(x), xlab="x", ylab="frequency") 
```

The above is sampled with *rbinom*.  In R we can also invoke the theoretical density function *dbinom*.
```{r}
# same using dbinom. ?dbinom
dbinom(0:10, size=n*(n-1)/2, prob=p)

# the distribution function Pr[X<=x]
pbinom(0:10, size=n*(n-1)/2, prob=p)

# fraction of no success (0) as function of size
1 - dbinom(0, size=100:110, prob=p)

n = 1:50
N = n*(n-1)/2           # N is a vector!!!
fa = 1 - dbinom(0,N,p)  # probability of success: 1-P[X=0]
# fa = pbinom(0,N,p, lower.tail=FALSE)  # P[X>0]

plot(n, fract, main="Birthday Problem", xlab="People", ylab="Fraction", col="red", pch=21, bg="grey",lwd=2) # as before
points(n, fa, type="l", lwd=2)      # analytically using dbinom
#?points
legend("bottomright",inset=0.02, legend=c("simulation", "analytic"), 
       pch=c(21,NA), pt.bg=c("grey",NA), lty=c(NA,1), col=c("red","black"))
abline(v=24)
```

Again we see that, for $n>23$, we expect that $>50\%$ of parties have matching birthdays.


# Did you understand?
* sample with replacement
* queries on vectors: sum, sort, duplicated entires, any
* matrix from vector
* apply more efficient than loops
* writing and invoking scripts (source)
* sampling (simple) vs analytical treatment (fast)
* binomial distribution and others: rbinom, dbinom, pbinom