---
title: "Simulated Data: The Birthday Problem"
author: "T. Manke"
date: "25 Feb 2016"
output: html_document
---

# The Problem:
Given $n$ people at a birthday party, how likely is it that at least two of them share a birthday?

# Sampling Excerices
Sample uniformly (with replacement) from 365 days (no Feb 29)

```{r}
n=40
bds=sample(1:365, n, replace=TRUE)
```

Are there any duplicated entries?

```{r}
dups=duplicated(bds)
```

What sort of object is dups?

```{r}
str(dups)
typeof(dups)
summary(dups)
sort(bds)
bds[dups]
sum(dups,na.rm=TRUE)
any(dups)
anyDuplicated(bds)
```

# Simulation in R
With those ingedients, determine the expected fraction of $m=1000$ parties with one or more duplicated birthdays.
```{r}
m=1000
bds=sample(365, n*m, replace=TRUE)
A=matrix(bds,m,n,byrow=TRUE)
x=apply(A,1,anyDuplicated)
fract=sum(x>0)/m
fract
```

Now plot the fraction against the size of the party $n \in [1,50]$
```{r}
fract=vector("numeric",50)
for (n in 1:50) {
  A=matrix(sample(365,n*m,replace=TRUE), m,n,byrow=TRUE)
  fract[n]=sum(apply(A,1,anyDuplicated)>0)/m
}
plot(1:50, fract, main="Birthday Problem: numerical simulation", xlab="People", ylab="Fraction", 
     type="o", col="red", pch=21, bg="grey",lwd=2)
#?plot
abline(h=0.5, lty=3, col="blue", lwd=2)
```

With $n>23$ people we expect more than half of all parties to have matching birthdays.

# Analytical treatment
For any given pair of people the probability of success (sharing a birthday) is small $p=1/365 \approx 0.0027$, but in the party setting 
there are many such comparisons $N=n(n-1)/2 \propto n^2$. The total number of successes can be modelled by the **binomial distribution** with success probability $p$ and $N$ number of trials.
Rather than simulating the process (above), we now draw $m$ random numbers from a binomial distribution
```{r}
m=100
n=24
p=1/365
x=rbinom(m,size=n*(n-1)/2, prob=p)
head(x)
table(x)
plot(x)
barplot(table(x), xlab="x", ylab="frequency")
plot(ecdf(x))

# understand dbinom
dbinom(0:10, size=n*(n-1)/2, prob=p)

n = 1:50
N = n*(n-1)/2
#N is a vector!!!
fa = 1 - dbinom(0,N,p)

plot(1:50, fract, main="Birthday Problem", xlab="People", ylab="Fraction", col="red", pch=21, bg="grey",lwd=2)
points(n, fa, type="l", lwd=2)
#?points
legend("bottomright",inset=0.02, legend=c("simulation", "analytic"), 
       pch=c(21,NA), pt.bg=c("grey",NA), lty=c(NA,1), col=c("red","black"))
abline(v=24)
```

Again we see that, for $n>23$, we expect that $>50\%$ of parties have matching birthdays.

# Multiple Testing correction
Now, let's consider statsitical testing. Statsitical tests are often phrased in terms of a null hypothesis
and data is used to calculate a score (for which the distribution is known).
Take again the binomial example. We may want to test the null hypothesis that a coin is fair ($p=0.5$)
and we have $N=10$ observations that resulted in 9 heads.
```{r}

# null hypothesis: fair coin ==> binomial with p=0.5
p=0.5
# observed data (independent) H H L H H H .... => 10 throws, 9 heads
# statistic 
N=10
S=9

# probability of observing 9 or more heads if coin is fair
# Notice that (S>=9) = (S>8)
pvalue=pbinom(S-1,N,p,lower.tail=FALSE)
pvalue

# try with S=8
# reject ?
```

In this case we would reject the null hypothesis of a fair coin.
Now let's test 10000 **fair coins**, each with 10 throws and a criterion (S>8) to reject the null hypothesis. How many rejections (false positives)?
```{r}
m=10000
x=rbinom(m,10,p)
fp=sum(x>8)
fp
```

Lets look more carefully at the  
```{r}
table(x)
# adjust breakpoints, histogram is for continuous distributions
hist(x,50,col="red")
abline(v=8)
```

Plot two distribution: one for number of successes, one for associated p-values
```{r}
pvs=dbinom(x,10,p)
par(mfcol=c(1,2))
barplot(table(x), space=0.5)
hist(pvs)
```

# Did you understand?
* sample with replacement
* queries on vectors: sum, sort, duplicated entires, any
* matrix from vector
* apply more efficient than loops
* plotting options, points, legend, abline
* analytical formalism more efficient than simulation
* hist, ecdf
* plot layout: par()
* binomial distribution and others: rbinom, dbinom, pbinom
* multiple testing correction