---
title: "Tidying a mess of data"
author: "Devon P. Ryan"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Initial setup

Before continuing, please run the following:

```{r install, echo=TRUE, eval=FALSE}
install.packages(c("readr", "tidyr", "dplyr", "ggplot2", "devtools"))
```

# Real data is very messy

- Typically not in a usable format
- Random empty values
- Poorly coded treatments

This **always** happens with other's datasets.

The process of correcting all of this is called "munging."

# With great power comes great annoyance

Congrats, you're now the R gal/guy. Everyone will now ask you for help and their data will be a total mess.

- Everyone likes doing crazy stuff in Excel
  - Missing values
  - Random coding, like "--" for missing
  - Multiple entries/column

| | Day | Escape Time | Distance Traveled | Quadrant Occupancy |
|-|:---:|:-----------:|:-----------------:|:------------------:|
| Mouse 1 | 1 | 60 | 100 | 45 / 27 |
| Mouse 2 | 1 | 50 | 110 | 45 / 27 |
| Mouse 3 | 1 | 52 | 00 | 45 / 27 |
| Mouse 1 | 2 | -- | -- | |

# The Hadleyverse

Much of modern data analysis relies heavily on packages from Hadley Wickham:

- readr
- tidyr
- reshape2
- dplyr (largely replaces reshape2)
- ggplot2 (the next section)

# readr

- Alternatives for `read.delim`, `read.csv`, etc.
- Handles remote files
- Consistent option names
- Only read some columns
- Error tolerant
- No automatic conversion of strings -> factors
- dplyr integration
- row names are never set

# readr example

```{r, echo=TRUE, warning=FALSE}
library("readr")
df = read_csv(col_types = "dd", col_names = c("x", "y"), skip = 1, "
1,2
a,b
")
```
```{r, echo=TRUE}
df
problems(df)
```

# readr column types

| Full name | Symbol | Example |
|:----------|:------:|--------:|
| `col_logical()` | l | `T`, `F`, `TRUE`, `FALSE` |
| `col_integer()` | i | 0, 22, 45 |
| `col_double()` | d | 3.1415926536 |
| `col_character()` | c | "I'm what's called a string!" |
| `col_skip()` | _ | |
| `col_number()` | n | "500.25€" |

> - N.B., there are more, see `vignette("column-types")`.

- Open the xlsx file in eXcel/Openoffice
  - Multiple levels of headers?!?!
  - Empty columns
  - Missing values
  - OMFG, there are commas instead of decimal points
    - Computers are Americans, they're easily confused by your strange foreign ways

The experiment:

- Two groups of mice
- 4/cage
- Food intake / weight measured over time
    
# readr example

File structure:

- Two header lines (ignore)
- 1 column of integers
- 18 columns of measurements (group 1)
- 1 blank column
- 17 columns of measurements (group 2)

```{r, echo=TRUE}
blah = locale("en", decimal_mark=",") # stupid commas
d = read_csv("../data/foodIntake.csv", skip=1, locale=blah)
head(d, n=1)
tail(d, n=1)
```

`readr` figured out the types and converted blanks to `NA` for us! Let's fix the first column names:

```{r}
colnames(d)[c(1,20)] = c("Week", "Skip")
```

# Tidy all the things

We loaded an ugly dataset, but how do we clean it.

**tidyr** and **dplyr**!

```{r, echo=TRUE}
library("tidyr")
library("dplyr", warn.conflicts = FALSE)
```

# tidyr/dplyr %>% 

**%>%** is called a "pipe"

- "Ctrl + shift + M" in RStudio
- From dplyr


Old school: **y = f(g(h(x)))**

- Hard to read from the inside out

New school: **x %>% h() %>% g() %>% f()**

>- More typing...
>- but more logical

# tidyr/dplyr %>% 

- **%>%** passes input as the first argument
    - **x %>% f(y)** is the same as **f(x, y)**
- This can be changed
    - **y %>% f(x, ., z)** is the same as **f(x, y, z)**

# tidyr gather()

Let's gather everything together so we have something like:

| Week | Cage | Group | Value |
|:----:|:----:|------:|------:|
| 1 | 528 | Control | 3.558 |
| 1 | 535 | Control | 3.775 |
| 1 | 525 | Treatment | 3.478 |

```{r, echo=TRUE}
d2 = d %>% select(-Skip) %>% gather(Cage, Value, -Week)
head(d2) # or "d2 %>% head()"
tail(d2)
```

Add the group

```{r, echo=TRUE}
d2$Group = c(rep(rep("Control", 33), 18), rep(rep("Treatment", 33), 17))
```

# Filtering

There are a bunch of `NA`s, which it'd be nice to remove. We can use `filter()` for that.

```{r, echo=TRUE}
d2 = d2 %>% filter(!is.na(Value))
tail(d2)
dim(d2)
```

# Group-based summarization

We'd like a group mean per week

```{r, echo=TRUE}
d3 = d2 %>% group_by(Week, Group) %>% summarise(Val=mean(Value))
head(d3)
```

# tidyr: Reshaping

We can do more than **`gather()`**

- spread()
- separate()
- unite()

```{r, echo=TRUE}
dem_polls = read_csv("http://elections.huffingtonpost.com/pollster/2016-iowa-presidential-democratic-caucus.csv")
```

```{r, echo=TRUE, eval=FALSE}
View(dem_polls)
```

Borrowed from: https://github.com/datasciencelabs/2016/blob/master/lectures/wrangling/data-wrangling-with-tidyr.Rmd

# tidyr: spread()

Sometimes you want a "wide-format" for saving as a file. This can be done with `spread()`

```{r, echo=TRUE}
dem_polls_gathered <- dem_polls %>% 
                         gather(key = candidate, value = percentage, 
                                Clinton:Undecided)
```
```{r, echo=TRUE, eval=FALSE}
View(dem_polls_gathered %>% spread(key = candidate, value = percentage))
```

# tidyr: separate()
     
Sometimes more than one value is present in each column. We can separate those:

```{r, echo=TRUE}
dem_polls_separate <- dem_polls %>% 
                        separate(col = `End Date`, into = c("y", "m", "d"))
```
```{r, echo=TRUE, eval=FALSE}
View(dem_polls_separate)
```

Note that the date is now split.

# tidyr: unite()

Of course if you can split columns, you can unite them too:

```{r, echo=TRUE}
dem_polls_united = dem_polls_separate %>% unite(col = end_date, y, m, d, sep = "/")
```
```{r, echo=TRUE, eval=FALSE}
View(dem_polls_united)
```
