---
title: "Manipulating datasets with dplyr"
author: "Devon P. Ryan"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Initial setup

If you've started a new session, please first do the following:

```{r}
library("readr")
library("tidyr")
library("dplyr", warn.conflicts = FALSE)

d = read_csv("../data/foodIntake2.csv", skip=1)
colnames(d)[c(1,20)] = c("Week", "Skip")
d2 = d %>% select(-Skip) %>% gather(Cage, Value, -Week)
d2$Group = c(rep(rep("Control", 33), 18), rep(rep("Treatment", 33), 17))
msleep = read_tsv("../data/msleep_ggplot2.txt")
```

This is the messy "food intake in two groups of mice over time" dataset from the tidyr section.

# What dplyr can do

- subset
    - columns
    - rows
- summarization
- Grouping
- Variable creation
- Combining datasets

# dplyr: Subset columns

It's common to want to temporarily want to ignore columns (e.g., excluding an outlier sample). **`select()`** can be used for this.

```{r}
d2 %>% select(Cage, Group) %>% head()
```

You can also exclude columns:

```{r}
d2 %>% select(-Cage, -Group) %>% head()
```

# dplyr: Fancy column selection

Specifying columns manually can get annoying.

```{r}
d2 %>% select(matches(".a."))  %>% head()
```

".a." means "match any name with an 'a' in the middle".

Other options:

- Range (by number or name)
- Starts with
- Ends with
- Contains a string

# dplyr: Subset rows

We did a bit of this previously in the tidyr section.

```{r}
head(d2)
d2 %>% filter(Group != "Control") %>% head()
```

Multiple filters can be applied

```{r}
head(d2)
d2 %>% filter(Group != "Control", Value > 4.0) %>% head()
```

There are other options than simply **`filter()`**

```{r}
d2 %>% select(Cage, Group) %>% distinct() %>% head()
```

- Sample a fraction of rows: **`sample_frac()`**
- Sample a number of rows: **`sample_n()`**
- Select rows by position: **`slice()`**
    - **`d2 %>% slice(10:20)`** == **`d2[c(10:20),]`**
- Select the top N rows after sorting: **`top_n()`**

```{r}
d2 %>% top_n(5, Value)
```

# dplyr: Summarizing data

Long tables are nice for plotting, but we often want means/medians/counts/etc. of columns.

```{r}
d2 %>% count(Cage, Group) %>% head()
d2 %>% filter(is.na(Value)) %>% count(Cage, Group) %>% head()
d2 %>% summarise(foo=n_distinct(Value))
d2 %>% filter(!is.na(Value)) %>% summarise(foo=n_distinct(Value))
d2 %>% filter(!is.na(Value)) %>% nrow() # or summarise(foo=n())
d2 %>% filter(!is.na(Value)) %>% summarise(avg=median(Value))
```

# Exercise

Going back to the `msleep_ggplot2.txt` file that we loaded earlier, what's the average time (in hours) that primates spend sleeping?












```{r}
msleep %>% filter(order == 'Primates') %>% summarise(ave=mean(sleep_total))
```

# dplyr: Grouping data

We can create group and then summarize within them.

```{r}
d2 %>% filter(!is.na(Value), Week < 10) %>% group_by(Group) %>% summarise(avg=mean(Value)) %>% head()
```

We can group by multiple things.

```{r, echo=TRUE}
d2 %>% filter(!is.na(Value)) %>% group_by(Group, Week) %>% summarise(avg=mean(Value)) %>% top_n(4, avg)
```

There are 8 results. Why? What happens if we swap the order of **Group** and **Week** in **`group_by`**?

# Exercise

Again, using the `msleep_ggplot2.txt` dataset, what is the average sleep time per order?









```{r}
msleep %>% group_by(order) %>% summarise(ave=mean(sleep_total))
```

# dplyr: Adding variables

It's often the case that we want to add on columns of summary variables. We can use **`mutate()`**.

```{r, echo=TRUE}
d2 %>% filter(!is.na(Value)) %>% mutate(blah=2*Value) %>% head()
```

We can alternatively drop the used columns with **`transmute()`**.

```{r, echo=TRUE}
d2 %>% filter(!is.na(Value)) %>% transmute(blah=2*Value) %>% head()
```

There are a lot of functions that could be applied.

```{r, echo=T}
d2 %>% filter(!is.na(Value)) %>% group_by(Cage) %>% mutate(blah=min_rank(desc(Value))) %>% head()
```

There are a bunch of functions that one can use. An interesting one is **`ntile`**, which is chunks things by value into **n** groups. Send the following through **`View()`** rather than **`head()`**.

```{r, echo=TRUE}
d2 %>% filter(!is.na(Value)) %>% group_by(Cage) %>% mutate(blah=ntile(Value, 3)) %>% head()
```

What's the per-tile median by group:

```{r, echo=TRUE}
d2 %>% filter(!is.na(Value)) %>% group_by(Cage) %>% mutate(blah=ntile(Value, 3)) %>% group_by(Group, blah) %>% summarise(avg=median(Value))
```

# dplyr: Combining datasets

Needing to combine datasets ends up being really really common. dplyr gives a **bunch** of options for that that are really fast.

A common use-case:

- We had two groups of mice
- Did some RNAseq
- Mapped the reads, got some counts, and ran things through DESeq2
- The output looks like this:

```{r}
tab = read_tsv("../data/DESeq2_example.txt")
head(tab[order(tab$padj),], n=3)
```

> - Raise your hand if you know which gene ENSMUSG00000037362 is.

We can use biomart to gene Ensembl gene IDs and their associated symbols/descriptions.

```{r, echo=TRUE}
anno = read_tsv("../data/DESeq2_annotation.txt")
```

Run **`head()`** or **`View()`** on these.

```{r, echo=TRUE}
tab2 = left_join(tab, anno, by=c("ENSEMBL" = "Ensembl Gene ID"))
head(tab2[order(tab2$padj),])
```

> - Ah, ENSMUSG00000037362 is Nov, a gene involved in wound healing/oncogenesis/etc.

There are a lot of different ways to combine datasets.

```{r, echo=TRUE}
t1 = data_frame(x1=c("A", "B", "C"), x2=c(1, 2, 3))
t2 = data_frame(x1=c("A", "B", "D"), x3=c(T, F, T))
```

- Mutating joins
- Filtering joins

# dplyr: Mutating joins - left_join

We've already seen **`left_join()`**

```{r, echo=TRUE}
left_join(t1, t2)
```

For a left join, the entire left table is output.

# dplyr: Mutating joins - right_join

If a left join joins to the left, guess what a right join does

```{r, echo=TRUE}
right_join(t1, t2)
```

# dplyr: Mutating joins - inner_join

Sometimes we only want the overlaps in the tables

```{r, echo=TRUE}
inner_join(t1, t2)
```

# dplyr: Mutating joins - full_join

But what if we want a single table with **everything**?

```{r, echo=TRUE}
full_join(t1, t2)
```

# dplyr: Combining datasets

You can get really fancy in your joins

```{r, echo=TRUE}
t3 = data_frame(a=rep(c("A", "B", "C"), 2),
                b=rep(c(1,2), each=3),
                c=runif(6))
t4 = data_frame(d=rep(c("A", "B", "C"), 2),
                e=rep(c(1,2), each=3),
                f=runif(6))
inner_join(t3, t4, by=c("a"="d", "b"="e"))
```

By default, joins occur on all columns with the same name across tables!
