---
title: "Correlation Coefficient"
author: "T. Manke"
date: "19 Feb 2016"
output: html_document
---

[//]: # Adopted from http://www-users.york.ac.uk/~mb55/meas/ba.pdf

# Simulating Data
Let's generate two numeric vectors (variables) of measurements: x1 and x2
```{r}
x1=1:100                          # vector
eps=rnorm(n=100, mean=-0.5, sd=5) # noise
x2 = x1 + eps 
```

We would like to assess the agreement between these two measurements.

# First attempt: visualize data
```{r}
plot(x1,x2)
smoothScatter(x1,x2)
```

We can add a regression line (linear model)

```{r, echo=FALSE}
cc=cor(x1,x2)
title=paste("cc=",round(cc,5))
plot(x1,x2,main=title)
abline(lm(x2~x1),col="red",lwd=2)
```

We may even test the significance. Null hypothesis: There is no linear relation between variables.
```{r}
cor.test(x1,x2)
```
Notice: if the variables come from same (or similar) experiements, the significance test is useless because we already know there will be a relation. Neither the CC nor the p-value can serve as a measure of agreement

```{r}
x3=rnorm(100,0,1)
x4=rnorm(100,0,1)
```

**Task**: How does these vectors look like? (--> plot,hist, summary)

# Scatterplot
```{r, echo=FALSE}
smoothScatter(x3,x4)
cor(x3,x4)
```

#Adding Outliers
```{r}
plot(c(x3,20),c(x4,100))
```

```{r}
cor(c(x3,20),c(x4,100), method="pearson")
cor(c(x3,20),c(x4,100), method="spearman")
```

Correlation Coefficient: 

* do not measure agreement
* single number *tries* to summarize whole correlation structure
* depends on method
* Spearman more robust against outliers
* significance is useless for many observations (most variables have $C \ne 0$)

**Homework**: Have a look at anscombe data set: ?anscombe