---
title: "P01: Data Exploration: Regulatory Modules from ChIP-seq"
author: "Thomas Manke"
date: '`r Sys.Date()`'
output: 
  html_document:
    code_folding: hide
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache=TRUE)
```

# Goal 
- Input:  genome-wide binding data for 113 transcription factors in yeast (ChIP-chip): --> 6000 x 113 matrix
- Analysis: identify combinations of TF
- Output: a reproducible analysis document (Rmd) and report (html)
- Reference: Lee et al. (2002) Transcriptional Regulatory Networks in Saccharomyces cerevisea.
- Data Set: chip="http://jura.wi.mit.edu/young_public/regulatory_network/binding_by_gene.tsv"

# Setup
Define libraries required for this project
```{r}
library(pheatmap)
```


# Skipped: Pre-processing
Steps: extract only p-values from full data, define proper names for rows, remove columns 
```{r preprocessing, eval=FALSE}
# Beware of slow connections ! 
chip="http://jura.wi.mit.edu/young_public/regulatory_network/binding_by_gene.tsv"
D=read.csv(chip, sep="\t", skip=1) # skip first line

### some filtering
rownames(D)=D[,1]                  # ORF-symbols in column 1, keep them as rownames 
ic=c(1:4,seq(6,230,by=2))          # define column indices to be excluded (only keep p-values)
P=D[,-ic]                          # new data-frame of p-values, exclude superfluous columns and ratios
####

# save cleaned data (only p-values) to file - for future use
# use write.table instead of write.csv - as I want to have tab-seperation
write.table(P, file="../data/chip.tsv", sep="\t", quote=FALSE) 
```

# Read Clean Data
Let's start with a cleaned-up data set. 

Now available at "https://raw.githubusercontent.com/maxplanck-ie/Rintro/master/data/chip.tsv"
(you might want to locate the *raw* file on https://github.com/maxplanck-ie/Rintro/).

Faster: have it local

```{r read}
fn="https://raw.githubusercontent.com/maxplanck-ie/Rintro/master/data/chip.tsv"  # remote filename
fn="../data/chip.tsv"                                                            # local: same but faster
P=read.csv(file=fn, sep="\t")
```


# Tasks: Sanity Checks
Inspect data. Below are some ideas.   
Try this iteratively on the console. So that you don't clutter your notebook.

```{r sanity, eval=FALSE}
str(P)             # structure
dim(P)             # dimensions
colnames(P)        # TF
anyNA(P)           # are there any undefined values
sum(P<0 | P>1)     # how many "strange" p-values --> errors?
```

**Query**: How many genes (actually: intergenic regions) are in this data frame?


## Summary Plots
```{r summary_plots}
boxplot(P)         # distribution
#hist(as.matrix(P)) # hist() requires matrix as input 
```

## Show me all the data
This might be slow.
```{r pheatmap, cache=TRUE}
pheatmap(P, show_rownames=FALSE)
```

# Transformations
```{r transform}
pt = 1e-3           # define P-value threshold
B = P < pt          # B is a Boolean matrix (FALSE/TRUE)
```

# Filter
Some genes (rows) have no associated TF. Some TF target no gene.
```{r filter}
nt = rowSums(B)     # number of transcription factors targetting a gene
ng = colSums(B)     # number of genes targetted by TF
D = B[nt>1, ng>40]  # define filtered data

```

**Query**: What is the dimension of the filtered matrix B 
```{r filtered_dim, eval=FALSE, echo=FALSE}
dim(B)
dim(D)
```

# Show me the processed data
```{r plot}
pheatmap(D+0, show_rownames=FALSE) # D + 0 is my trick to convert logical matrix to numeric matrix (FALSE=0, TRUE=1) 
```

# Main Results: All Correlations
```{r correlations}
C = cor(D)
pheatmap(C)
```

#  Explore specific factor combinations

```{r smoothScatter}
t1="FHL1"
t2="RAP1"
plot( P[,t1], P[,t2] )                     # not very meaningful
smoothScatter( log(P[,t1]), log(P[,t2]) )  # scatterplot of log(p-values) for two TF 
abline(h=log(pt), v=log(pt), col="red")    # show thresholds
```

Correlation Coefficient: `r C[t1,t2]`. 

# Contigency tables
```{r contTables}
table(B[,t1])               # 194 genes bound by t1="FHL1"
names(which(B[,t1]))        # their names

tb=table(B[,t1],B[,t2])     # contingency table: 119 genes bound by FHL1 and RAP1
tb

fisher.test(tb)             # a test at last: the overlap is highly unexpected
```

# Investigating Multiple Overlaps
upsetR is better (more general) than Venn diagrams
```{r upsetR}
library(UpSetR)
upset( as.data.frame(B+0), sets = c("FHL1", "RAP1", "MCM1", "STE12") )
```



# Review
* Data munging: typically requires basic processing and filtering steps ==> use clean data
* data exploration and visualization: heatmap(), smoothScatter()
* quantitative data descriptions: cor(), table()
* some tests: fisher.test()
* R notebooks: run code chunks interactively, or generate full reports
