---
title: "Data Exploration: Regulatory Modules from ChIP-seq"
author: "Thomas Manke"
date: "`r date()`"
output: html_document
---

## Obtain and filter data
Goal: Download data from one of the first large scale ChIP-studies from the Rick Young lab:
Lee et al. (2002) Transcriptional Regulatory Networks in Saccharomyces cerevisea.

Subgoals:
1. Find Rick Young lab homepage (google), go to the "Data Download"" section, and find the relevant publication.

2. In the "Download Raw Data" section copy the second link "binding of regulators to genes (text file)".
This file contains binding data (intensity ratios and their p-values) for more than 6000 yeast genes and 113 yeast transcription factors

3. filter data to obtain a matrix of p-values where (rows=genes) and (columns=TFs)

```{r, eval=FALSE}
# define a filename (notice that R can read from http)
# Beware: For slow network connections it will be faster to first download the file to local storage and work from there.

chip="http://jura.wi.mit.edu/young_public/regulatory_network/binding_by_gene.tsv"

###############################
# Data munging: Trial and errors have been commented out 
# first problem: two headers --> skip first --> skip line 1
# D=read.table(file=chip, header=TRUE,  skip=1, sep="\t")
# nrow(D)  # 511 << 6000 !!!???

# second problem: quote symbols in names (')
# cf=count.fields(chip,sep="\t") # many NA --> quoted fields got missing

D=read.table(chip, header=TRUE,  sep="\t", quote="\"", skip=1)
rownames(D)=D[,1]          # ORF-symbols in column 1, keep them as rownames 
ic=c(1:4,seq(6,230,by=2))  # define column indices to be excluded (only keep p-values)
P=D[,-ic]                  # new data-frame of p-values, exclude superfluous columns and ratios
###############################

# save cleaned data for future use
write.table(P,file="../data/chip.tsv", sep="\t", quote=FALSE)  # write p-values to file
```

### Cleaned Real
Let's start with a cleaned-up data set. Now available at "https://raw.githubusercontent.com/maxplanck-ie/Rintro/master/data/chip.tsv"
(locate this *raw* file on https://github.com/maxplanck-ie/Rintro/)

```{r}
fn="https://raw.githubusercontent.com/maxplanck-ie/Rintro/master/data/chip.tsv"  # remote filename
P=read.table(file=fn, header=TRUE, sep="\t")
```

Describe data and sanity checks
```{r, eval=FALSE}
str(P)            # describe the data structure
which(is.na(P))   # make sure everything is properly defined
which(P<0 || P>1) # make sure p-values are all between 0 and 1
```

Now we can start looking at something more interesting
```{r}
# defined boolean matrix of TF-gene interactions (based on p-value threshold)
pt=1e-3
B=P<pt  # result is matrix.

# reduce problem size for illustration purposes, since heatmap is slow 
i=rowSums(B)>1   # only genes with sufficient number of TF-binding
j=colSums(B)>40  # only TF with sufficient number of targets

# heatmap needs numeric matrix ==> convert logical to integer 
B=1*B  
# equivalent tricks: B[B]=1; mode(B) = "integer", 
# but _not_ as.integer(B)  nor as.numeric(B)  # this will strip attributes, i.e. dimensions
# B[]=as.numeric(B) works as it fills a vector into the prexisting matrix B

heatmap(B[i,j], scale="none", col=c("white","black"), labRow=FALSE)
#d3heatmap(B[i,j], scale="none", col=c("white","black"), distfun = function(c) dist(c, method="binary"))
```

Explore relation between FHL1 and RAP1
```{r}
smoothScatter(log(P[,"FHL1"]),log(P[,"RAP1"]),main="smoothScatter of p-values")  # scatterplot of log(p-values) for two TF 
abline(h=log(pt), v=log(pt), col="red")                                          # "arbitrary" thresholds
cor(B[,"FHL1"],B[,"RAP1"])                                                       # Phi-coefficient = Pearson CC for binary variables
```

Notice that some correlations may be undefined because all entries for certain TF (=columns) are zero
```{r}
cor(B[,"ABF1"], B[,"HAA1"]) 
sum(B[,"HAA1"])
```

Find matrix elements with maximal correlations (ignore NA's and ignore diagonal=1)
```{r}
C=cor(B)
m=max(C[upper.tri(C)], na.rm=TRUE)  # maximum of upper-triagonal matrix. Ignore NA for maximum
idx=which(C==m, arr.ind = TRUE)     # return matrix indices. There are two index pairs  because C is symmetric.
idx
i1=idx[1,1]
j1=idx[1,2]

cat(colnames(C)[c(i1,j1)], C[i1,j1])
```

Contigency tables
```{r}
table(B[,"FHL1"])               # 194 genes bound by FHL1
tb=table(B[,"FHL1"],B[,"RAP1"]) # contingency table: 119 genes bound by FHL1 and RAP1
tb

fisher.test(tb)                 # the overlap is highly unexpected
chisq.test(tb)
```

Generalize: Find other groups of TF with common targets 
```{r}
C=cor(B[,j])                             # correlation matrix. Alternative: C=cor(log(P+eps))
heatmap(C,col = rev(grey.colors(100)) ) 
```

### Packages
Heatmaps can summarize overall data, but are often difficult to explore.
Additional tools are needed. Enter the world of *R packages*.

Explore CRAN "https://cran.r-project.org" orGoogle "interactive heatmaps in R"
```{r}
#install.packages("heatmaply")   # only need to install once (but takes some time: ~10min)
library(heatmaply)               # only need once per session. notice messages and warnings
heatmaply(C,col = rev(grey.colors(100)) ) 
```

Investigating overlaps:
```{r}
# Which genes are commonly bound by FHL1 and RAP1?
names(which(B[,"FHL1"]==1 & B[,"RAP1"]==1))

# create a Venn Diagram of target gene overlaps (with package gplots)
library(gplots)
L=list(FHL1=which(B[,"FHL1"]==1),RAP1=which(B[,"RAP1"]==1),MCM1=which(B[,"MCM1"]==1), STE12=which(B[,"STE12"]==1) )
vn=venn(L)

# better than Venn
library(UpSetR)
upset( as.data.frame(B), sets = c("FHL1", "RAP1", "MCM1", "STE12"), empty.intersections = TRUE )
```

***

# Review:
* Data munging: import often requires basic processing and filtering ==> use clean data
* data retyping: data.frame --> Boolean matrix --> Integer
* data exploration and visualization heatmap(), smoothScatter()
* quantitative data descriptions: cor(), table()
* tests: fisher.test(), chi2.test()
* packages: heatmaply, gplots (venn), UpSetR
