---
title: "Luciferase - An Example"
author: "Thomas Manke"
date: '`r Sys.Date()`'
output:
  html_document:
    code_folding: show
    toc: yes
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache=TRUE)
```

# Read the data
```{r read}
# notice: use first column as row.names
D=read.delim2("../data/luciferase.tsv", row.names=1)
D
```

# Transform
Barplot (from base R) will need matrix. So need to convert data.frame --> matrix. 
At the same time log-transform might be useful --> avoid gaps
```{r transform}
M=as.matrix(log(D))
```

# Barplot
These are first attempts to generate boxplots with base R.
They will not be very meaningful, but they can highlight some problems
```{r barplot_with_baseR}
palette(rainbow(4,alpha=0.7)) # set some palette for color mapping

barplot(M, beside = TRUE, col=1:4)
legend("topleft", legend = rownames(M),  fill = 1:4)

# Taking the transpose (t(M) below) may be somwhat closer to the desired output
cols=rep(1:4, each=3)  # define colours
sp=c(0.2, 1)           # control the gap between the bars
barplot(t(M), beside = TRUE, col=cols, space = sp)
legend("topleft", legend = colnames(M),  fill = cols)
# Notice that the colour assignment assumes a specific order an implicit understanding of the data
```

# Problems
- no easy way to plot error bars on barplots (discouraged) --> boxplot
- no easy way to plot gap on axis --> other packages (thank you google): plotrix:gap.barplot --> better solution: log-transform
- no way for R to know groups (relation between columns): only humans take interpretation of column names --> colour choice
- need to define some representative value for groups (mean) and variation around (standard deviation)

Unfortunately the data is not "tidy" :-(

# Tidy the data
The tidyverse way. Very powerful, but it forces the adoption of some grammar rules

```{r tidy}
library(tidyverse)  # this package (and dependencies) will have to be installed before

# now keep the Treatment as column (c.f. the first efforts)

D=read.delim2("../data/luciferase.tsv") 
```

## Reformat the data
Observation: We have two categorical variables (Treatment, Condition), 1 measurement (number)

Goal: Gather all columns and 

- collect them as categorical variable ("Cond"),
- call the observed value = "number"
- keep variable "Treatment"

We will do this in a pipeline (%>%)

```{r gather}
D %>% gather(Cond, number, -Treatment) %>% head  # just print the head for now (might be a big data set)
```

Now expand the above pipe to deal with the Condition variable: unify Cond.
Remove the suffix from string (mutate variable)
```{r mutate}
# thankfully this is easy (just remove suffix .1 and .2) --> mutate and use regexp
D = D %>% gather(Cond, number, -Treatment) %>% mutate(Cond=str_remove(Cond,".\\d"))
```


# Boxplot
This is the prefered way, as you will show the distribution in a more meaningful way.

```{r boxplot}
ggplot(D, aes(x=Cond, y=number, colour=Cond)) +
  geom_boxplot() +                              # a boxplot summary
  geom_jitter() +                               # add actual data points
  scale_y_continuous(trans = 'log2') +          # on a logarithmic scale
  facet_grid(~Treatment)  +                     # "facet" over treatments
  theme(axis.text.x = element_text(angle = 90)) # use theme to change axis labels/angles
```

# Barplot
If you really have to use a barplot. Barplots are frequently used, but discouraged

https://www.data-to-viz.com/caveat/error_bar.html

```{r barplot}
# First define some mean number (mn) and standard deviation (nd) for each combination (Treatment, Cond)
Db = D %>% group_by(Treatment, Cond) %>% summarize(nm=mean(number), nd=sd(number))

ggplot(Db, aes(x=Cond, y=nm, fill=Cond)) +
  geom_bar(stat="identity") +                         # plots barplot
  geom_errorbar(aes(ymin=nm-nd, ymax=nm+nd)) +        # adds errorbar
  scale_y_continuous(trans = 'log2') +                # log. scale
  facet_grid(~Treatment) +                          
  theme(axis.text.x = element_text(angle = 90))       # use theme() to change axis labels/angles
```