---
title: "Nested Models & Logistic Regression"
author: "Thomas Manke"
date: "12 Feb 2018"
output: html_document
---

# Fitting and comparing (nested) models
```{r}
fit1=lm(Petal.Width ~ Petal.Length, data=iris)            # fit a linear model
fit2=lm(Petal.Width ~ Petal.Length + Species, data=iris)  # add one variable Species
anova(fit1, fit2)
```

*Interpretation*: adding variable "Species" significantly improves fit of Model 2.

# Limitations of correlation and regression analysis
**Homework**: Look at the Anscombe data set with 4 different variable pairs. Calculate Spearman and Pearson correlation coefficients for the same 4 pairs.
```{r, eval=FALSE}
?anscombe
str(anscombe)
head(anscombe)
example(anscombe)
?cor
```

# Logistic regression
Goal: predict species based on features
```{r}
x=iris
x$setosa = (x$Species=="setosa")    # create new binary variable
x$col="black"                       # default color for plot
plot(setosa ~ Sepal.Length, data=x, col=x$col)

model=glm(setosa ~ Sepal.Length, data=x, family=binomial(logit))     # generalized linear model --> logistic regression

points(fitted(model) ~ Sepal.Length, data=x, col="red", pch=20)

x$col="black"
x$col[fitted(model)>0.5]="red"     # classify (and color red) if setosa is probable
plot(setosa ~ Sepal.Length, data=x, col=x$col, main="Classification and Misclassification")
```

```{r}
x$Species = NULL                                       # remove Species variable from x
model=glm(setosa ~ ., family=binomial(logit), data=x)  # model with _all_ features/variables

x$col="black"
x$col[fitted(model)>0.5]="red"
plot(setosa ~ Sepal.Length, data=x, col=x$col)         # perfect predictor
```


***
# Review:
* comparing different model, assess improvement
* limitations of correlation and regression analysis
* logistic regression: generalized linear model glm()
