---
title: "Review"
author: "Thomas Manke"
date:  "`r date() `"
output: html_document
---
# 2020-04-16
## Data Selection
```{r}
sc=c(1,3)                      # select column by index
sr=grep("^v", iris$Species)    # select rows by 
iris[,sc] 
iris[sr,]
```

## Formulas
```{r}
boxplot( Petal.Length ~ Species, data = iris)
plot( Petal.Length ~ Petal.Width, data = iris)
```


# 2020-04-15
## Structured Data = Easy Analysis & Plotting
Many powerful functions. Many customizable default parameters
```{r , eval=FALSE}
x=iris$Petal.Length  # numeric vector
summary(x)
plot(x)
hist(x)
boxplot(x)
```

## Palette: Mapping Numbers to Colours
```{r base_palettes}
palette("default")               # optional: default
palette()
barplot(1:20,col=1:20)

palette(rainbow(18,alpha=0.5))   # explain this command
palette()
barplot(1:20,col=1:20)

palette(heat.colors(18))
barplot(1:20,col=1:20)
palette("default")                # reset to default
```

## Use predefined palettes. 

Other packages have even more palettes: `library(RColorBrewer)`

## Own colour palettes 
Define your own colours - if you must
```{r own_palettes}
alpha=seq(0.1, 1.0, length.out = 50)   # control transparency
mc=rgb(0,0,1,alpha)                    # my colours
mc
plot(iris$Petal.Length,col=mc)


```


# 2020-04-14
## Functions

- Everything is a function: $y = f(x_1,x_2, \ldots, p_1, \ldots)$
- *structured* inputs: variables ($x_i$) and parameters ($p_i \to$ defaults?)
- *structured* outputs: $y$

Examples:
```{r}
sqrt(2)
sin(pi)
1+2
```
## Variables

- creation
- access: by index (`v[3]`), by name (`v["name"]`)
- recycling: `1:10 + c(1,2)`

Examples:
```{r variables}
# variable creation
x=exp(2)
v=sqrt(1:5)
names(v)=letters[1:5]

# variable access
v
v[3]    # by index
v["d"]  # by name
v[v<2]  # by logic

1:10 + c(1,2) # recycle
```


## Warning and errors

```{r warnings, eval=FALSE}
1:10 + c(1,2,3)
sqrt(-1:2)
```

```{r errors, eval=FALSE}
sqrt("a")
```


